Modelling multi-scale state-switching functional data with hidden Markov models

### 2021-05-22 Report/Sections/results.tex
update reviewer 1, comments 1-4a

Understanding animal behaviour is important for conservation efforts, as environmental changes caused by anthropogenic activity can directly impact animal behaviour \citep{Sutherland:1998}. HMMs have been used to understand how diving behaviours of various species are affected by disturbances (e.g., \citet{DeRuiter:2017} and \citet{Isojunno:2017}). For killer whales, we are interested in categorizing different diving behaviours to identify potential foraging dives. Northern resident killer whales feed almost exclusively on calorie-rich Chinook salmon (\textit{Oncorhynchus tshawytscha}) \citep{Ford:2006}, which typically occur deeper and are less numerous than smaller types of salmon \citep{Ford:2009}. Northern resident killer whales therefore must expend significant amounts of energy to capture Chinook \citep{Williams:2009,Noren:2011,Wright:2017}. 
%It is therefore of interest to determine the energetic expenditure of killer whales hunting for Chinook as well as the frequency at which they catch prey, especially since many Chinook salmon in the area are either threatened or endangered \citep{Ford:2015}. 
Acceleration data can be used to estimate an animal's energy expenditure \citep{Green:2009,Wilson:2019}, but the animal's behavioural state must be accounted for in order to obtain accurate estimates \citep{Dot:2016}. Therefore, understanding both the behavioural state of the killer whale and the distribution of acceleration within each behavioural state is needed to determine the true energetic requirements of the animal.

@@

The data we use were collected on September 2, 2019 from 12:49 pm to 6:06 pm PDT and consist of depth and acceleration over time. Observations were collected at a rate of 50 Hz using a CATs biologger (Customizable Animal Tracking Solutions, {\em{www.cats.is}}). Acceleration was measured in three dimensions, which together represent the complete range of movement of an animal (forward/backward, upward/downward, and right/left). Tri-axial acceleration readings are common in these types of tags and are often used to infer animal behaviour such as foraging \citep{Cade:2017,Fehlmann:2017,Wright:2017}. The act of attaching and detaching the tag caused anomalous behaviour before 1:20 pm and after 6:00 pm, so observations taken during these time periods are ignored. There were also periods of time when the tag failed to record observations, resulting in data gaps between 2:25 pm and 2:37 pm and between 4:07 pm and 5:07 pm. To preprocess the data, we smooth the depth and acceleration curves by taking a moving average within a window of $1/10^{th}$ of a second. We then define a killer whale ``dive" as any continuous interval of data that occurs below 0.5 meters in depth and lasts for at least 10 seconds. Data are preprocessed in part with the \textit{divebomb} package in Python \citep{Nunes:2018}. The preprocessed data contain a total of 267 dives, all of which are displayed in Figure \ref{fig:data}. Each dive is treated as one curve, and the sequence of dives makes up the coarse-scale process. Specifically, the observed coarse-scale observations $y = \big\{y_1,\ldots,y_{267}\big\}$ make up a sequence of dive durations in seconds. For dive $t$, the fine-scale observations are contained in $y^*_{t} = \big\{y^*_{t,1},\ldots,y^*_{t,T^*_t} \big\}$, which is a sequence containing the within-dive acceleration data in units of meters per second squared. The collection of all acceleration data is denoted as $y^* = \big\{y^*_1,\ldots,y^*_{267}\big\}$.

@@

Defining a suitable model to describe this killer whale kinematic data involves selecting an appropriate number of hidden states, model structure, and emission distributions for both the coarse- and fine-scale observations.

@@

We then select the specific emission distribution of $\Z_{t,\tilde t^*}$. First, we assume that $\Ztwo_{t,\tilde t^*}$ and all three components of $\Zone_{t,\tilde t^*}$ are independent of one another when conditioned on the dive types and subdive states. To reduce model complexity, we also assume that the three sets of fine-scale emission parameters are shared across the two dive types $\left(\theta^{*(1,i^*)} = \theta^{*(2,i^*)} \equiv \theta^{*(\cdot,i^*)} \text{ for } i^* = 1,2,3\right)$. This implies that the subdive states within dive type 1 have the same interpretation as those within dive type 2.
To specify the emission distribution of $\Zone_{t,\tilde t^*}$, consider the sequence $\{\Zone_{t,1},\ldots,\Zone_{t,T^*_t}\}$ for a particular dive $t$. We assume that each of the three components of this sequence are Normally distributed as in Equation (\ref{eqn:carhmm}) and all components are independent of one another when conditioned on the subdive states $\tilde X_{t,1},\ldots,\tilde X_{t,\tilde T^*_t}$. Each component is assumed to have its own mean and variance parameters, but all components share the same auto-correlation parameter. Thus the distribution of $\Zone_{t,\tilde t^*}$ given $X^*_{t,\tilde t^*} = i^*$ has parameters $\mathbf{\mu}_A^{*(\cdot,i^*)} \in \mathbb{R}^3$, $\mathbf{\sigma}_A^{*(\cdot,i^*)} \in \mathbb{R}^3$, and $\phi_A^{*(\cdot,i^*)} \in [0,1]$.
To specify the emission distribution of $\Ztwo_{t,\tilde t^*}$, we assume that given $\tilde X^*_{t,\tilde t^*} = i^*$, $\Ztwo_{t,\tilde t^*}$ follows a Gamma distribution parameterized by its mean $\mu_W^{*(\cdot,i^*)}$ and standard deviation $\sigma_W^{*(\cdot,i^*)}$. In addition, $\Ztwo_{t,1},\ldots,\Ztwo_{t,T^*_t}$ are assumed to be independent of one another given the subdive state sequence $\big\{\tilde X_{t,1}, \ldots, \tilde X_{t,T_t^*}\big\}$. We do not include $\Ztwo_{t,\tilde t^*-1}$ in the distribution of $\Ztwo_{t,\tilde t^*}$ because the auto-correlation evident from the lag plot is not severe and may be explained by subsequent observations occurring within the same subdive state. 
%
%The final model is a hierarchical hidden Markov model with a traditional HMM on the coarse scale. Each fine-scale sequence $y^*_t$ is transformed to $\z$ using Equation (\ref{eqn:z}) and $\Z_{t,\tilde t^*}$ is modelled with a CarHMM.
%We assume $\Zone_{t,\tilde t^*}$ is Normally distributed with auto-correlation while $\Ztwo_{t,\tilde t^*}$ is Gamma distributed without auto-correlation. We also assume that $\Ztwo_{t,\tilde t^*}$ and all components of $\Zone_{t,\tilde t^*}$ are each independent of one another when conditioned on the subdive state. The emission distribution of each subdive state does not depend upon the dive type.

@@

In addition to the CarHHMM-DFT, we consider three variations for comparison. As in the full model, each of the following models assume that all components of $\Z_{t,\tilde t^*}$ are conditionally independent of one another given the dive types and subdive states:
\begin{enumerate}
    \item An \textbf{HHMM-DFT}, which models the coarse-scale observations with an HMM and transforms the fine-scale observations using Equation (\ref{eqn:z}), but models $\Z_{t,\tilde t^*}$ as emissions from a simple HMM rather than a CarHMM.
    \item A \textbf{CarHHMM}, which models the coarse-scale observations with an HMM, transforms the fine-scale observations using Equation (\ref{eqn:z}), and models $\Zone_{t,\tilde t^*}$ as emissions of a CarHMM. However, the ``wiggliness"  $\Ztwo_{t,\tilde t^*}$ is omitted from this model altogether.
    \item A \textbf{CarHMM-DFT}, which models the coarse-scale observations as an independent and identically distributed sequence of dives, transforms the fine-scale observations using Equation (\ref{eqn:z}), and models $\Z_{t,\tilde t^*}$ as emissions of a CarHMM. This model assumes that there is only one dive type.
    %, so the fine-scale probability transition matrix $\Gamma^*$ is the same for every dive. 
\end{enumerate}
%
Each of the three candidate models above leaves out one important aspect of the full CarHHMM-DFT: the HHMM-DFT assumes there is no auto-correlation between fine-scale observations, the CarHHMM does not incorporate ``wiggliness" $\big(\Ztwo_{t,\tilde t^*}\big)$, and the CarHMM-DFT lacks a hierarchical structure and thus does not distinguish between dive types.

@@

\[\hat \delta^{*(1)} = \begin{pmatrix} 0.087 & 0.731 & 0.182 \end{pmatrix}, \enspace \text{and} \enspace \hat \delta^{*(2)} = \begin{pmatrix} 0.401 & 0.496 & 0.103 \end{pmatrix}\]
%
for dive types 1 and 2.
%The estimated coarse-scale stationary distribution, $\hat{\delta}$, indicates that 
In summary, about 79\% of dives are short dives of type 1, and
%The estimated coarse-scale transition probability matrix $\hat \Gamma$ further indicates that 
the whale performs an average of 4.72 short type 1 dives before switching to dive type 2 and an average of 1.24 longer type 2 dives before switching back to dive type 1. This finding is consistent with those of \citet{Tennessen:2019b} and \citet{Williams:2009}, both of whom describe common bouts of short resting dives before a killer whale performs a longer, more energy-intensive deep dive. %The fine-scale stationary distributions $\hat{\delta}^{*(i^*)}$ indicate that 
Further, this killer whale is in the less active subdive state 1 40\% of the time during a dive of type 1 compared to only only 9\% of the time during a dive of type 2. Less active swimming behaviour is consistent with the need for marine mammals to conserve energy when diving to depth and holding their breath for long periods \citep{Williams:1999,Hastie:2006}. Figure \ref{fig:labeled_dives} shows the decoded dive behaviour of six selected dives, and Section 1.3 of the supplementary material also shows the probability of each dive type and subdive state given the data and the fitted model.

>>>>

Understanding animal behaviour is important for conservation efforts, as environmental changes caused by anthropogenic activity can directly impact animal behaviour \citep{Sutherland:1998}. HMMs have been used to understand how diving behaviours of various species are affected by disturbances (e.g., \citet{DeRuiter:2017} and \citet{Isojunno:2017}). For killer whales, we are interested in categorizing different diving behaviours to identify potential foraging dives. Northern resident killer whales feed almost exclusively on calorie-rich Chinook salmon (\textit{Oncorhynchus tshawytscha}) \citep{Ford:2006}, which typically occur deeper and are less numerous than smaller types of salmon \citep{Ford:2009}. Northern resident killer whales therefore must expend significant amounts of energy to capture Chinook \citep{Williams:2009,Noren:2011,Wright:2017}. 
Acceleration data can be used to estimate an animal's energy expenditure \citep{Green:2009,Wilson:2019}, but the animal's behavioural state must be accounted for in order to obtain accurate estimates \citep{Dot:2016}. Therefore, understanding both the behavioural state of the killer whale and the distribution of acceleration within each behavioural state is needed to determine the true energetic requirements of the animal.

@@

The data we use were collected on September 2, 2019 from 12:49 pm to 6:06 pm PDT and consist of depth and acceleration over time. Observations were collected at a rate of 50 Hz using a CATs biologger (Customizable Animal Tracking Solutions, {\em{www.cats.is}}). Acceleration was measured in three dimensions, which together represent the complete range of movement of an animal (forward/backward, upward/downward, and right/left). Tri-axial acceleration readings are common in these types of tags and are often used to infer animal behaviour such as foraging \citep{Cade:2017,Fehlmann:2017,Wright:2017}. The act of attaching and detaching the tag caused anomalous behaviour before 1:20 pm and after 6:00 pm, so observations taken during these time periods are ignored. There were also periods of time when the tag failed to record observations, resulting in data gaps between 2:25 pm and 2:37 pm and between 4:07 pm and 5:07 pm. To preprocess the data, we smooth the depth and acceleration curves by taking a moving average within a window of $1/10^{th}$ of a second. We then define a killer whale ``dive" as any continuous interval of data that occurs below 0.5 meters in depth and lasts for at least 10 seconds. Data are preprocessed in part with the \textit{divebomb} package in Python \citep{Nunes:2018}. The preprocessed data contain a total of 267 dives, all of which are displayed in Figure \ref{fig:data}. Each dive is treated as one curve, and the sequence of dives makes up the coarse-scale process. Specifically, the observed coarse-scale observations $y = \big\{y_1,\ldots,y_{267}\big\}$ make up a sequence of dive durations in seconds, and the coarse-scale hidden states $x = \big\{x_1,\ldots,x_{267}\big\}$ represent the corresponding dive types. For dive $t$, the fine-scale observations are contained in $y^*_{t} = \big\{y^*_{t,1},\ldots,y^*_{t,T^*_t} \big\}$, which is a sequence containing the within-dive acceleration data in units of meters per second squared. The fine-scale hidden states of dive $t$, $x^*_{t} = \big\{x^*_{t,1},\ldots,x^*_{t,T^*_t} \big\}$, represent the subdive behavioural states of the killer whale. The collection of all acceleration data is denoted as $y^* = \big\{y^*_1,\ldots,y^*_{267}\big\}$ and the collection of all unknown subdive behaviours is denoted as $x^* = \big\{x^*_1,\ldots,x^*_{267}\big\}$.

@@

The primary goal of this case study is to jointly estimate the dive types and (more importantly) the subdive behaviours of this killer whale, so only hierarchical models were considered to describe the kinematic data. Defining a suitable hierarchical model involves selecting an appropriate number of hidden states, model structure, and emission distributions for both the coarse- and fine-scale observations.

@@

We then select the specific emission distributions of $\Z_{t,\tilde t^*}$ for all dive types and subdive states. First, we assume that $\Ztwo_{t,\tilde t^*}$ and all three components of $\Zone_{t,\tilde t^*}$ are independent of one another when conditioned on the dive types and subdive states. To reduce model complexity, we also assume that the three sets of fine-scale emission parameters are shared across the two dive types $\left(\theta^{*(1,i^*)} = \theta^{*(2,i^*)} \equiv \theta^{*(\cdot,i^*)} \text{ for } i^* = 1,2,3\right)$. This implies that the subdive states within dive type 1 have the same interpretation as those within dive type 2.
To specify the emission distribution of $\Zone_{t,\tilde t^*}$, consider the sequence $\{\Zone_{t,1},\ldots,\Zone_{t,T^*_t}\}$ for a particular dive $t$. We assume that each of the three components of this sequence are Normally distributed as in Equation (\ref{eqn:carhmm}) and all components are independent of one another when conditioned on the subdive states $\tilde X_{t,1},\ldots,\tilde X_{t,\tilde T^*_t}$. Each component is assumed to have its own mean and variance parameters, but all components share the same auto-correlation parameter. Thus the distribution of $\Zone_{t,\tilde t^*}$ given $X^*_{t,\tilde t^*} = i^*$ has parameters $\mathbf{\mu}_A^{*(\cdot,i^*)} \in \mathbb{R}^3$, $\mathbf{\sigma}_A^{*(\cdot,i^*)} \in \mathbb{R}^3$, and $\phi_A^{*(\cdot,i^*)} \in [0,1]$.
To specify the emission distribution of $\Ztwo_{t,\tilde t^*}$, we assume that given $\tilde X^*_{t,\tilde t^*} = i^*$, $\Ztwo_{t,\tilde t^*}$ follows a Gamma distribution parameterized by its mean $\mu_W^{*(\cdot,i^*)}$ and standard deviation $\sigma_W^{*(\cdot,i^*)}$. In addition, $\Ztwo_{t,1},\ldots,\Ztwo_{t,T^*_t}$ are assumed to be independent of one another given the subdive state sequence $\big\{\tilde X_{t,1}, \ldots, \tilde X_{t,T_t^*}\big\}$. We do not include $\Ztwo_{t,\tilde t^*-1}$ in the distribution of $\Ztwo_{t,\tilde t^*}$ because the auto-correlation evident from the lag plot is not severe and may be explained by subsequent observations occurring within the same subdive state.

@@

In addition to the CarHHMM-DFT, we consider three variations for comparison. As in the full model, each of the following models assume that all components of $\Z_{t,\tilde t^*}$ are conditionally independent of one another given the dive types and subdive states:
\begin{enumerate}
    \item An \textbf{HHMM-DFT}, which models the coarse-scale observations with an HMM and transforms the fine-scale observations using Equation (\ref{eqn:z}), but models $\Z_{t,\tilde t^*}$ as emissions from a simple HMM rather than a CarHMM.
    \item A \textbf{CarHHMM}, which models the coarse-scale observations with an HMM, transforms the fine-scale observations using Equation (\ref{eqn:z}), and models $\Zone_{t,\tilde t^*}$ as emissions of a CarHMM. However, the ``wiggliness"  $\Ztwo_{t,\tilde t^*}$ is omitted from this model altogether.
    \item A \textbf{CarHMM-DFT}, which models the coarse-scale observations as an independent and identically distributed sequence of dives, transforms the fine-scale observations using Equation (\ref{eqn:z}), and models $\Z_{t,\tilde t^*}$ as emissions of a CarHMM. This model assumes that there is only one dive type.
\end{enumerate}
%
Each of the three candidate models above leaves out one important aspect of the full CarHHMM-DFT: the HHMM-DFT assumes there is no auto-correlation between fine-scale observations, the CarHHMM does not incorporate ``wiggliness" $\big(\Ztwo_{t,\tilde t^*}\big)$, and the CarHMM-DFT lacks a hierarchical structure and thus does not distinguish between dive types.

@@

\[\hat \delta^{*(1)} = \begin{pmatrix} 0.087 & 0.731 & 0.182 \end{pmatrix}, \enspace \text{and} \enspace \hat \delta^{*(2)} = \begin{pmatrix} 0.401 & 0.496 & 0.103 \end{pmatrix}\]
%
for dive types 1 and 2.
In summary, about 79\% of dives are short dives of type 1, and the whale performs an average of 4.72 short type 1 dives before switching to dive type 2 and an average of 1.24 longer type 2 dives before switching back to dive type 1. This finding is consistent with those of \citet{Tennessen:2019b} and \citet{Williams:2009}, both of whom describe common bouts of short resting dives before a killer whale performs a longer, more energy-intensive deep dive.
Further, this killer whale is in the less active subdive state 1 40\% of the time during a dive of type 1 compared to only only 9\% of the time during a dive of type 2. Less active swimming behaviour is consistent with the need for marine mammals to conserve energy when diving to depth and holding their breath for long periods \citep{Williams:1999,Hastie:2006}. Figure \ref{fig:labeled_dives} shows the decoded dive behaviour of six selected dives, and Section 1.3 of the supplementary material also shows the probability of each dive type and subdive state given the data and the fitted model.
